---
title: "Which groups are at highest risk for low uptake of flu vaccine?"
author: Chris Okitondo and Irene Cavros
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
editor: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

# Summary/Abstract

[\*\*To be completed (PART 5)]{.underline}

# Introduction

### General Background Information

[\*\*To be completed (PART 4)]{.underline}

### Questions/Hypotheses to be addressed

Using the aforementioned database, our research question of interest is
as follows:

Which groups are at highest risk for low uptake of flu vaccine?

# Methods

### Description of data and data source

The data source used for this project is the National Health Interview
Survey (NHIS.) It is a harmonized set of data beginning in 1963 with
information on general health status, acute and chronic illness,
functional limitations, access to care, insurance coverage, and health
behaviors for the U.S. population. On average, the survey covers 100,000
persons in 45,000 households each year. PUMS NHIS currently contains
over 17,000 integrated variables from 1963 to present.

As the outcome of interest, researchers focused on those who, in this
household survey between the years of 2012 and 2021, answered no when
they were asked whether they had ANY flu vaccine in the past 12 months.

Variables assessed are as follows:

1\. Demographics: age, sex, race, ethnicity, veteran status

2\. Socioeconomic status: education level, total combined family income,
employment status, paid sick leave as part of job benefits, ratio of
family income to poverty threshold

3\. General health: perceived health status, categorical BMI

4\. Conditions: ever told had asthma, ever told had cancer, ever told
had coronary heart disease

5\. Health behaviors: alcohol drinking status, cigarette smoking status

6\. Access to care: has usual place for medical care, medical care
delayed due to cost (past 12 months)

7\. Health insurance: health insurance coverage status

8\. Mental health: anxiety level, depression level

#### Study population

The raw data for our period of interest directly extracted from the NHIS
IPUMS survey initially had had 792,916 observations. After processing
the raw data and removing all observations with missing data in any of
our predictors of interest, 106980 individuals were included in our
analysis. Table 1 below provides a summary of demographic
characteristics of our study population.

[\*\*To be completed (PART 4)]{.underline}

### Statistical analysis

*Explain anything related to your statistical analyses.*

[\*\*To be completed (PART 4)]{.underline}

# Results

### Exploratory/Descriptive analysis

\*\* Table 1: Demographic characteristics of the study population \*\*

### Loading Table 1 from the Exploratory Analysis

```{r, warning=FALSE, message=FALSE}
result_table1 <- readRDS(here("results", "table1_demographics.rds"))
result_table1
```

\*\*

### Full analysis

##### Model 1: Focusing on only demogrphics characteristics using the flu vaccine uptake as our outcome

```{r, warning=FALSE, message=FALSE}
result_table2 <- readRDS(here("results", "demographics_model.rds"))
result_table2
```

### Model 2: Using all our predictors of interest in the model with flu vaccine uptake as our outcome

```{r, warning=FALSE}
result_table3 <- readRDS(here("results", "full_model.rds"))
result_table3
```

\*\*

# Discussion

### Summary and Interpretation

*Summarize what you did, what you found and what it means.*

[\*\*To be completed (PART 5)]{.underline}

### Strengths and Limitations

*Discuss what you perceive as strengths and limitations of your
analysis.*

[\*\*To be completed (PART 5)]{.underline}

### Conclusions

*What are the main take-home messages?*

[\*\*To be completed (PART 5)]{.underline}

# References

*Include citations in your Rmd file using bibtex, the list of references
will automatically be placed at the end*

[\*\*To be completed (PART 5)]{.underline}
